Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-20 paper/ms.tex

\dfmtodo{Make this less douchey:}
In the current era of large-scale time domain astronomy, scalable methods for
probabilistic inference are of great interest.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials but it does not require evenly
spaced observations or homoscedastic noise.
This form of the covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that this function can also be a sufficient effective model in many other
cases.
We present a mathematical description of the method, details of the
implementation, and compare it to existing scalable Gaussian Process methods.
We demonstrate the application of this method by applying it to simulated and
real astronomical time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method is flexible, fast, and most importantly, interpretable.
There are a wide range of potential applications within astronomical data
analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, simple, and fast.

@@

For a GP kernel to be valid, it must produce a positive definite covariance
matrix for all input coordinates.
For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
Fourier transform of a positive finite measure.
This means that the power spectrum of a positive definite kernel must be
positive for all frequencies.

@@

The power spectrum is computed from taking the square of the Fourier transform
of the data time series; hence, the power-spectrum must be non-negative.  In
constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
then it is possible for the power spectrum to go negative, which violates the
definition of the power spectrum.
Consequently, if any of these coefficients is negative, it is necessary to
check that the power spectrum is still non-negative.  Note that a non-negative
power-spectrum does not require the auto-correlation
function to be positive for all $\tau$.  The positive power spectrum is
related to the positive eigenvalues of the covariance matrix which are
necessary for a positive-definite matrix in limiting cases
\citep{Messerschmitt:2006}.  We find empirically that requiring an everywhere
positive power spectrum results in positive eigenvalues for the covariance
matrix, and so we describe here how to ensure a positive power spectrum using
Sturm's theorem.
%Relation of positive power spectrum to positive definite covariance matrix.
%Power spectrum of a sum of kernels.

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that scale well with amount of data.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the application of this method by applying it to simulated and
real astronomical time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
The method is flexible, fast, and most importantly, interpretable.
There are a wide range of potential applications within astronomical data
analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, simple, and fast.

@@

For a GP kernel to be valid, it must produce a positive definite covariance
matrix for all input coordinates.
For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
Fourier transform of a positive finite measure.
This means that the power spectrum of a positive definite kernel must be
positive for all frequencies.  This seems true intuitively:  since the
power spectrum is the square of the Fourier transform of the data time
series, it must be non-negative.

@@

%The general case with multiple \celerite\ terms can be checked using Sturm's
%theorem. The power spectrum is computed from taking the square of the Fourier 
%transform of the data time series; hence, the power-spectrum must be non-negative.  
%In constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
%then it is possible for the power spectrum to go negative, which violates the
%definition of the power spectrum.
%Consequently, if any of these coefficients is negative, it is necessary to
%check that the power spectrum is still non-negative.  Note that a non-negative
%power-spectrum does not require the auto-correlation
function to be positive for all $\tau$.  %The positive power spectrum is
%related to the positive eigenvalues of the covariance matrix which are
%necessary for a positive-definite matrix in limiting cases
%\citep{Messerschmitt:2006}.  %We find empirically that requiring an everywhere
%positive power spectrum results in positive eigenvalues for the covariance
%matrix, and so we describe here how to ensure a positive power spectrum using
%Sturm's theorem.
%Relation of positive power spectrum to positive definite covariance matrix.
%Power spectrum of a sum of kernels.
